---
title: "Body size and cranial shape differentiation in urban and rural house mice (Mus musculus domesticus)" 
output: html_notebook
#Biorxiv doi: https://doi.org/10.64898/2026.05.16.725634
---

## Read Data

```{r}

library(dplyr)
S1 <- readxl::read_excel("Stables_Data.xlsx", sheet = 2, skip = 1, n_max = 97)
colnames(S1)
S1$Metro <- S1$"Metro Region"
S1$ID <- S1$`Collector ID`

skull  <- S1[S1[, grep("in the skull\r\ndataset",  colnames(S1))[1]] == "yes", ]
weight <- S1[S1[, grep("in the weight\r\ndataset", colnames(S1))[1]] == "yes", ]
linear <- S1[S1[, grep("in the linear\r\ndataset", colnames(S1))[1]] == "yes", ]

clean_nas <- function(df) {
  df <- as.data.frame(lapply(df, function(x) {
    if (is.character(x)) x[x == "N/A"] <- NA
    x
  }), stringsAsFactors = FALSE)
  return(df)
}

skull <- clean_nas(skull)
weight <- clean_nas(weight)
linear <- clean_nas(linear)

cat("Skull n =", nrow(skull), "\n")
cat("Weight n =", nrow(weight), "\n")
cat("Linear n =", nrow(linear), "\n")

```

## Load in the necessary packages

```{r}
packages <- c("ggplot2", "geomorph", "Morpho", "ggpattern", "car", "cowplot", 
              "ggsignif", "dplyr", "paran", "emmeans", "ggforce", "patchwork", 
              "Rvcg", "writexl")

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

library(ggplot2)
library(geomorph)
library(Morpho)
library(ggpattern)
library(car)
library(cowplot)
library(ggsignif)
library(dplyr)
library(paran)
library(emmeans)
library(ggforce)
library(patchwork)
library(Rvcg)
library(writexl)

R.version.string
packageVersion("geomorph")
packageVersion("Morpho")
packageVersion("car")
packageVersion("paran")
```

## Sample

```{r}
dim(skull)
table(skull$Habitat)
table(skull$Metro)
print(skull$ID)
```

### Define landmark columns

```{r}

# Total: 84 Landmarks: MSP: 1 -- 10 -> (10) RLP: 11 -- 27 -> (17) LLP: 28 -- 44 -> (17) RZC: 45 -- 64 > (20) LZC: 65 -- 84 -> (20)

landmark_cols <- grep("^[XYZ]\\d+$", colnames(skull))
landmark_data <- apply(as.matrix(skull[, landmark_cols]), 2, as.numeric)

# Create a raw coordinate array
coords <- arrayspecs(landmark_data, p = 84, k = 3)
dimnames(coords)[[3]] <- skull$ID

# Export coords with ID
write_xlsx(cbind(ID = skull$ID, as.data.frame(landmark_data)), "coords.xlsx")
```

### Define sliding semi-landmark curves

```{r}
# Define semi-landmark curves
RZC_semilandmarks <- define.sliders(landmarks = 45:64, nsliders = 18)
LZC_semilandmarks <- define.sliders(landmarks = 65:84, nsliders = 18)

semi.lm <- rbind(RZC_semilandmarks, LZC_semilandmarks)
```

# 1. Generalized Procrustes Analysis (GPA)

```{r}

# Perform GPA with sliding semi-landmarks 
gpa_results <- gpagen(coords, curves = semi.lm, ProcD = TRUE, print.progress = TRUE)
print(gpa_results)

# Extract Centroid Size
aligned_coords <- gpa_results$coords   # Procrustes aligned coords
centroid <- (gpa_results$Csize)         # centroid size
log_centroid <- log(gpa_results$Csize)  # logarithm of centroid size

dim(aligned_coords)  # Landmarks × 3 × specimens (72)
length(log_centroid)  # equal to number of specimens (72)
length(centroid)      # eequal to number of specimens (72)
nrow(skull)

# GPA should return 48 fixed landmarks & 36 semilandmarks (sliders) - there are 40 total points across two semi-landmark curves, but the first and last point of each curve is designated as a 'fixed' point (total = 4)
```

### Geomorph Dataframe

```{r}

gdf <- geomorph.data.frame(
  aligned_coords = aligned_coords,
  centroid     = as.numeric(centroid), # raw centroid
  log_centroid = as.numeric(log_centroid), #log centroid
  ID           = skull$ID,
  Habitat      = skull$Habitat,
  Metro        = skull$Metro,
  Sex          = skull$Sex,
  femur        = as.numeric(skull$Femur.Length..mm.),
  volume       = as.numeric(skull$Endocast.Volume..mm3.),
  total        = as.numeric(skull$Total.Length..mm.),
  tail         = as.numeric(skull$Tail.Length..mm.),
  BL           = as.numeric(skull$Body.Length..mm.),
  weight       = as.numeric(skull$Weight..g.)
)

print(gdf$log_centroid)
```

# 2. Procrustes Permutation Anova (procD.lm)

```{r}
proc <- procD.lm(aligned_coords ~ Metro * Habitat * Sex + centroid, data = gdf, SS = "II")
summary(proc)
print(proc)

# tests for normality (Q-Q)
plot(proc)
residuals_proc <- residuals(proc)
  
standardized_residuals <- scale(residuals_proc)
qqnorm(standardized_residuals, main = "Q-Q Plot of Standardized Residuals")
qqline(standardized_residuals, col = "red", lwd = 2)

# Export output (SUPPLEMENTARY TABLE S7)
proc_summary <- as.data.frame(summary(proc)$table)
proc_summary$Term <- rownames(proc_summary)
proc_summary <- proc_summary[, c("Term", setdiff(names(proc_summary), "Term"))]  
write_xlsx(proc_summary, "SUP_TABLE_S7.xlsx")

```

# 3. Principal Component Analysis

```{r}
pca_result <- gm.prcomp(aligned_coords)
names(pca_result)
names(summary(pca_result))
head(summary(pca_result))
pca_scores <- pca_result$x
dim(pca_scores)
summary(pca_result)
colnames(pca_scores)

# Bind skull and PC scores in a new object, skull_pca
pc_df <- as.data.frame(pca_scores)
colnames(pc_df) <- paste0("PC", 1:ncol(pc_df))
skull_pca <- cbind(skull, pc_df)

# Export Output (SUPPLEMENTARY TABLE S3)
pc_df$ID <- skull$ID
pc_df$Specimen <- skull$Specimen
pc_df <- pc_df[, c("ID", paste0("PC", 1:ncol(pca_scores)))]
write_xlsx(pc_df, "SUP_TABLE_S3.xlsx")

# Export Output (SUPPLEMENTARY TABLE S4)
pca_summary <- as.data.frame(t(summary(pca_result)$PC.summary))
pca_summary$Eigenvalues <- pca_result$sdev^2
pca_summary$Component <- rownames(pca_summary)
pca_summary <- pca_summary[, c("Component", "Eigenvalues", setdiff(names(pca_summary), c("Component", "Eigenvalues")))]
colnames(pca_summary) <- c("Component", "Eigenvalues", "Variance Percent", "Cumulative Percent")
write_xlsx(pca_summary, "SUP_TABLE_S4.xlsx")
```

## Parallel Analysis

```{r}
aligned <- aligned_coords
n <- dim(aligned)[3]
p <- dim(aligned)[1]
k <- dim(aligned)[2]

# Flatten each specimen into a vector (p*k)
flat_coords <- t(apply(aligned, 3, function(x) as.vector(x)))

set.seed(1)
pa_result <- paran(flat_coords, iterations = 1000, graph = TRUE, centile = 0, quietly = FALSE,
                   n = n - 1)

pa_export <- data.frame(
  Component      = 1:(n - 1),
  Unadjusted     = pa_result$Ev[1:(n - 1)],
  Estimated_Bias = pa_result$Ev[1:(n - 1)] - pa_result$AdjEv[1:(n - 1)],
  Adjusted       = pa_result$AdjEv[1:(n - 1)],
  Retained       = pa_result$AdjEv[1:(n - 1)] > 1
)

# Export (SUPPLEMENTARY TABLE S5)
write_xlsx(pa_export, "SUP_TABLE_S5.xlsx")
```

# 4. Multivariate Analysis of Covariance (MANCOVA)

```{r}
num_pcs <- 9
mancova_data <- data.frame(
  pca_scores[, 1:num_pcs],   # PCA scores
  Sex      = skull$Sex,      
  Metro      = skull$Metro,      
  Habitat      = skull$Habitat,      
  centroid = centroid
)

mancova_result <- manova(as.matrix(mancova_data[, 1:num_pcs]) ~  Metro * Habitat * Sex + centroid, data = mancova_data)
summary(mancova_result)

# Export Output
names(summary(mancova_result))
mancova_summary <- as.data.frame(summary(mancova_result)$stats)
mancova_summary$Term <- rownames(mancova_summary)
mancova_summary <- mancova_summary[, c("Term", setdiff(names(mancova_summary), "Term"))]
write_xlsx(mancova_summary, "SUP_TABLE_S8.xlsx")
```

# 5. Allometry

```{r}
# plotAllometry
colors <- c("Urban" = "#C4773B", "Rural" = "#3B6E8F")

allometry_model_1 <- procD.lm(aligned_coords ~ log_centroid * Habitat, data = gdf, iter = 999, SS = "II")
plotAllometry(allometry_model_1, size = gdf$log_centroid, pch = 19,
              col = colors[gdf$Habitat], predLine = TRUE)
recorded_1 <- recordPlot()
summary(allometry_model_1)

allometry_model_2 <- procD.lm(aligned_coords ~ log_centroid + Habitat, data = gdf, iter = 999, SS = "II")
plotAllometry(allometry_model_2, size = gdf$log_centroid, pch = 19,
              col = colors[gdf$Habitat], predLine = TRUE)
recorded_2 <- recordPlot()
summary(allometry_model_2)

pdf("allometry_model_1.pdf", width = 88/25.4, height = 88/25.4)
replayPlot(recorded_1)
dev.off()

pdf("allometry_model_2.pdf", width = 88/25.4, height = 88/25.4)
replayPlot(recorded_2)
dev.off()

# CAC
# Perform CAC decomposition (Common Allometric Component)
cac_out <- CAC(x = aligned_coords, size = log_centroid, log = FALSE)
CAC_scores <- as.numeric(cac_out$CACscores) #Extract CAC scores

# Perform a correlation test (Pearson's R) between the CAC and PC1
PC1_scores <- pca_scores[, "Comp1"]
Pearson_r_CAC.PC1 <- cor.test(CAC_scores, PC1_scores) 
Pearson_r_CAC.PC1
```

# 6. Visualizations

### Find Reference Specimen

```{r}
# findMeanSpec function - finds specimen in the dataset whose landmark coordinates most closely compare to the mean established by GPA. 

Ref_Spec <- findMeanSpec(aligned_coords) 
print(Ref_Spec)
Ref_Index <- Ref_Spec   
Ref_ID <- gdf$ID[Ref_Index]
Ref_ID #MHB_096

print(gdf$ID)
gdf$ID[61]

ref_index <- which(gdf$ID == Ref_ID)
cat("Reference specimen:", Ref_ID, "(", as.character(gdf$ID[Ref_Index]), ")\n")
```

## Habitat - Mean Shapes & Heatmaps

```{r}
# Load in the reference mesh
reference_mesh <- vcgPlyRead("MHB_096_FNL_MESH.ply", updateNormals = TRUE)

# Extract GPA coordinates for reference specimen
coords_ref_gpa <- gpa_results$coords[,, Ref_Index]

# Calculate overall mean shape for the entire dataset
overall_mean_gpa <- mshape(gpa_results$coords)

# Warp reference mesh to the overall mean shape
cat("Warping reference mesh to overall mean shape...\n")
mean_mesh <- warpRefMesh(
  mesh = reference_mesh,
  mesh.coord = coords_ref_gpa,
  ref = overall_mean_gpa,
  color = NULL,
  centered = TRUE
)

# Calculate group mean shapes (urban and rural)
urban_mean_gpa <- mshape(gpa_results$coords[,, gdf$Habitat == "Urban"])
rural_mean_gpa <- mshape(gpa_results$coords[,, gdf$Habitat == "Rural"])

# Magnification factor
mag <- 20

# Warp urban
cat("Warping reference mesh to urban mean...\n")
urban_mesh <- plotRefToTarget(
  M1 = urban_mean_gpa,
  M2 = overall_mean_gpa,
  mesh = mean_mesh,
  method = "surface",
  mag = mag
)

# Warp rural
cat("Warping reference mesh to rural mean...\n")
rural_mesh <- plotRefToTarget(
  M1 = rural_mean_gpa,
  M2 = overall_mean_gpa,
  mesh = mean_mesh,
  method = "surface",
  mag = mag
)

# Remove material colors and save
mean_mesh$material$color <- NULL
urban_mesh$material$color <- NULL
rural_mesh$material$color <- NULL

vcgPlyWrite(mean_mesh, "AAA_overall_mean_mesh.ply")
vcgPlyWrite(urban_mesh, "AAA_urban_mean_mesh_mag_20.ply")
vcgPlyWrite(rural_mesh, "AAA_rural_mean_mesh_mag_20.ply")
cat("Complete\n")


# Urban-to-Mean heatmap
heatmap_urban <- meshDist(
  urban_mesh,
  overall_mesh,
  plot = FALSE,
  rampcolors = c("blue", "cyan", "yellow", "orange", "red"),
  sign = FALSE
)

vcgPlyWrite(heatmap_urban$colMesh, "AAA_HEATMAP_urban_to_mean_20.ply")

# Rural-to-mean heatmap
heatmap_rural <- meshDist(
  rural_mesh,
  overall_mesh,
  plot = FALSE,
  rampcolors = c("blue", "cyan", "yellow", "orange", "red"),
  sign = FALSE
)

vcgPlyWrite(heatmap_rural$colMesh, "AAA_HEATMAP_rural_to_mean_20.ply")

cat("Complete\n")
```

## Habitat - Residual Mean Shapes, (CAC removed), and heatmaps

```{r}
# Define groups
group <- gdf$Habitat
urban_index <- which(group == "Urban")
rural_index <- which(group == "Rural")

# Perform the CAC decomposition (Common Allometric Component)
cac_out <- CAC(x = aligned_coords, size = centroid, log = TRUE)

# Back-project all residual shapes into Procrustes space using restoreShapes
overall_mean_gpa <- mshape(aligned_coords) 
resid_shapes <- restoreShapes(cac_out$RSCscores, cac_out$RSC, overall_mean_gpa) 
# (restoreShapes returns a 3D array: p x 3 x n) the object 'resid_shapes' is now all of the shapes in the GPA coordinate system, but represents CAC removed shapes.

# Compute means from resid_shapes using mshape
resid_mean_RSC  <- mshape(resid_shapes)                      # p x 3 (residual mean)
urban_mean_RSC  <- mshape(resid_shapes[,, urban_index])      # p x 3 (urban residual mean)
rural_mean_RSC  <- mshape(resid_shapes[,, rural_index])      # p x 3 (rural residual mean)

# Create the residual mean mesh by warping the reference mesh to resid_mean_RSC using warpRefMesh
      
cat("Warping RSC mean reference mesh ...\n")
    mesh_resid_mean <- warpRefMesh( 
        mesh = reference_mesh,
        mesh.coord = coords_ref_gpa,
        ref = resid_mean_RSC,
        color = NULL,
        centered = TRUE)
        
# warp resid_mean_mesh to the group means (Habitat: Urban and Rural) using plotRefToTarget.
    
cat("Warping RSC mean to urban RSC mean...\n")    
mesh_resid_urban <- plotRefToTarget(
        M1 = urban_mean_RSC,     
        M2 = resid_mean_RSC,    
        mesh = mesh_resid_mean,  
        method = "surface",
        mag = mag)

cat("Warping RSC mean to rural RSC mean...\n")  
      mesh_resid_rural <- plotRefToTarget(
        M1 = rural_mean_RSC,
        M2 = resid_mean_RSC,
        mesh = mesh_resid_mean,
        method = "surface",
        mag = mag)

# Save  
mesh_resid_mean$material$color <- NULL
mesh_resid_urban$material$color <- NULL
mesh_resid_rural$material$color <- NULL

vcgPlyWrite(mesh_resid_mean, "AAB_CAC_removed_mean_mesh.ply")
vcgPlyWrite(mesh_resid_urban, "AAB_CAC_removed_urban_mean_mesh.ply")
vcgPlyWrite(mesh_resid_rural, "AAB_CAC_removed_rural_mean_mesh.ply")

heatmap_urban_resid <- meshDist(
  mesh_resid_urban,
  mesh_resid_mean,
  plot = FALSE,
  rampcolors = c("blue", "cyan", "yellow", "orange", "red"),
  sign = FALSE
)

vcgPlyWrite(heatmap_urban_resid$colMesh, "AAB_HEATMAP_UrbanVsMean_Resid_20.ply")

heatmap_rural_resid <- meshDist(
  mesh_resid_rural,
  mesh_resid_mean,
  plot = FALSE,
  rampcolors = c("blue", "cyan", "yellow", "orange", "red"),
  sign = FALSE
)

vcgPlyWrite(heatmap_rural_resid$colMesh, "AAB_HEATMAP_RuralVsMean_Resid_20.ply")

cat("Complete\n")
```

# 7. GLMs

```{r}
weight$Weight <- as.numeric(weight$Weight..g.)
weight$BL     <- as.numeric(weight$Body.Length..mm.)
weight$BC     <- as.numeric(weight$Body.Condition..g.mm.)


colnames(linear)
linear$femur  <- as.numeric(linear$Femur.Length..mm.)
linear$BL     <- as.numeric(linear$Body.Length..mm.)
linear$ML     <- as.numeric(linear$Femoral.Mediolateral.Width..mm.)

colnames(skull)
skull$centroid <- as.numeric(gdf$centroid)
skull$BL       <- as.numeric(skull$Body.Length..mm.)
skull$femur    <- as.numeric(skull$Femur.Length..mm.)
skull$volume   <- as.numeric(skull$Endocast.Volume..mm3.)


cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                              Weight (g)                          ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
Weightlm <- lm(Weight ~ Metro * Sex * Habitat, data = weight)
shapiro.test(residuals(Weightlm))
leveneTest(Weight ~ Metro * Sex * Habitat, data = weight)
WeightAnova <- Anova(Weightlm, type = "II")
print(WeightAnova)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                           Weight (g) + BL                        ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
Weightlm2 <- lm(Weight ~ Metro * Sex * Habitat + BL, data = weight)
shapiro.test(residuals(Weightlm2))
WeightAnova2 <- Anova(Weightlm2, type = "II")
print(WeightAnova2)
emmWeightlm2 <- emmeans(Weightlm2, ~ Habitat)
pairs(emmWeightlm2, adjust = "tukey")
print(emmWeightlm2)

emm_weight <- as.data.frame(emmWeightlm2)
rural_weight <- emm_weight$emmean[emm_weight$Habitat == "Rural"]
urban_weight <- emm_weight$emmean[emm_weight$Habitat == "Urban"]
((rural_weight - urban_weight) / rural_weight) * 100  # Urban % lighter

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                       Body Condition (g/mm)                      ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
BClm <- lm(BC ~ Metro * Sex * Habitat, data = weight)
shapiro.test(residuals(BClm))
leveneTest(BC ~ Metro * Sex * Habitat, data = weight)
BCAnova <- Anova(BClm, type = "II")
print(BCAnova)
emmBClm <- emmeans(BClm, ~ Habitat)
pairs(emmBClm, adjust = "tukey")
print(emmBClm)

emm_BC <- as.data.frame(emmBClm)
rural_BC <- emm_BC$emmean[emm_BC$Habitat == "Rural"]
urban_BC <- emm_BC$emmean[emm_BC$Habitat == "Urban"]
((rural_BC - urban_BC) / rural_BC) * 100  # Urban % lower BC than rural


cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                       Femur Length (mm)                          ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
femurlm <- lm(femur ~ Metro * Sex * Habitat, data = linear)
shapiro.test(residuals(femurlm))
leveneTest(femur ~ Metro * Sex * Habitat, data = linear)
femurAnova <- Anova(femurlm, type = "II")
print(femurAnova)
emmfemurlm <- emmeans(femurlm, ~ Habitat)
pairs(emmfemurlm, adjust = "tukey")
print(emmfemurlm)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                       Body Length (mm)                           ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
BLlm <- lm(BL ~ Metro * Sex * Habitat, data = linear)
shapiro.test(residuals(BLlm))
leveneTest(BL ~ Metro * Sex * Habitat, data = linear)
BLAnova <- Anova(BLlm, type = "II")
print(BLAnova)
emmBLlm <- emmeans(BLlm, ~ Habitat)
pairs(emmBLlm, adjust = "tukey")
print(emmBLlm)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                       Mediolateral Width (mm)                    ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
MLlm <- lm(ML ~ Metro * Sex * Habitat, data = linear)
shapiro.test(residuals(MLlm))
leveneTest(ML ~ Metro * Sex * Habitat, data = linear)
MLAnova <- Anova(MLlm, type = "II")
print(MLAnova)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                 Mediolateral Width + femur (mm)                  ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
MLlm2 <- lm(ML ~ Metro * Sex * Habitat + femur, data = linear)
shapiro.test(residuals(MLlm2))
MLAnova2 <- Anova(MLlm2, type = "II")
print(MLAnova2)
emmMLlm2_sex     <- emmeans(MLlm2, ~ Sex)
emmMLlm2_hab     <- emmeans(MLlm2, ~ Habitat)
pairs(emmMLlm2_sex,     adjust = "tukey"); print(emmMLlm2_sex)
pairs(emmMLlm2_hab,     adjust = "tukey"); print(emmMLlm2_hab)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                    Centroid Size                                 ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
centroidlm <- lm(centroid ~ Metro * Sex * Habitat, data = skull)
shapiro.test(residuals(centroidlm))
leveneTest(centroid ~ Metro * Sex * Habitat, data = skull)
centAnova <- Anova(centroidlm, type = "II")
print(centAnova)
emmcentroidlm <- emmeans(centroidlm, ~ Habitat)
pairs(emmcentroidlm, adjust = "tukey")
print(emmcentroidlm)

emm_cent <- as.data.frame(emmcentroidlm)
rural_cent <- emm_cent$emmean[emm_cent$Habitat == "Rural"]
urban_cent <- emm_cent$emmean[emm_cent$Habitat == "Urban"]
((rural_cent - urban_cent) / urban_cent) * 100  # Rural % larger centroid than urban

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                    Centroid Size  + BL                           ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
centroidlm2 <- lm(centroid ~ Metro * Sex * Habitat + BL, data = skull)
shapiro.test(residuals(centroidlm2))
centAnova2 <- Anova(centroidlm2, type = "II")
print(centAnova2)
emmcentroidlm2 <- emmeans(centroidlm2, ~ Habitat)
pairs(emmcentroidlm2, adjust = "tukey")
print(emmcentroidlm2)


cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                    Endocast Volume (mm³)                         ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
volumelm <- lm(volume ~ Metro * Sex * Habitat, data = skull)
shapiro.test(residuals(volumelm))
leveneTest(volume ~ Metro * Sex * Habitat, data = skull)
volAnova <- Anova(volumelm, type = "II")
print(volAnova)
emmvollm <- emmeans(volumelm, ~ Habitat)
pairs(emmvollm, adjust = "tukey")
print(emmvollm)

cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║                    Endocast Volume (mm³) + Centroid              ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n")
volumelm2 <- lm(volume ~ Metro * Sex * Habitat + centroid, data = skull)
shapiro.test(residuals(volumelm2))
volAnova2 <- Anova(volumelm2, type = "II")
print(volAnova2)
emmvollm2_hab   <- emmeans(volumelm2, ~ Habitat)
emmvollm2_sex   <- emmeans(volumelm2, ~ Habitat | Sex)
emmvollm2_metro <- emmeans(volumelm2, ~ Habitat | Metro)
pairs(emmvollm2_hab,   adjust = "tukey"); print(emmvollm2_hab)
pairs(emmvollm2_sex,   adjust = "tukey"); print(emmvollm2_sex)
pairs(emmvollm2_metro, adjust = "tukey"); print(emmvollm2_metro)


cat("\n\n")
cat("╔══════════════════════════════════════════════════════════════════╗\n")
cat("║          t.test (Welch's two-sample) For PC1 ~ Habitat           ║\n")
cat("╚══════════════════════════════════════════════════════════════════╝\n\n") 
t_result <- t.test(pca_scores[, 1] ~ skull_pca$Habitat)
t_result


## Export Output
anova_list <- list(
  "Weight"                       = WeightAnova,
  "Weight + BL"                  = WeightAnova2,
  "Body Condition (BC)"          = BCAnova,
  "Femur"                        = femurAnova,
  "Body Length (BL)"             = BLAnova,
  "Mediolateral width (ML)"      = MLAnova,
  "Mediolateral width + femur"   = MLAnova2,
  "Centroid"                     = centAnova,
  "Centroid + BL"                = centAnova2,
  "Endocast Volume"              = volAnova,
  "Endocast Volume + Centroid"   = volAnova2
)

combined <- do.call(rbind, lapply(names(anova_list), function(name) {
  a <- as.data.frame(anova_list[[name]])
  a$Term <- rownames(a)
  a$Model <- name
  a[, c("Model", "Term", setdiff(names(a), c("Model", "Term")))]
}))

write_xlsx(combined, "SUP_TABLE_S6.xlsx")
```

# 8. Figures

## Parameters

```{r}
habitat_colors <- c("Urban" = "#C4773B", "Rural" = "#3B6E8F")
metro_shapes   <- c("New York City" = 21, "Philadelphia" = 22, "Richmond" = 24)

# Re-ordering factors such that urban appears on the left of the plot
weight$Habitat    <- factor(weight$Habitat,    levels = c("Urban", "Rural"))
linear$Habitat    <- factor(linear$Habitat,    levels = c("Urban", "Rural"))
skull$Habitat     <- factor(skull$Habitat,     levels = c("Urban", "Rural"))
skull_pca$Habitat <- factor(skull_pca$Habitat, levels = c("Urban", "Rural"))

plot_theme <- theme_minimal(base_family = "Arial", base_size = 12) +
  theme(
    legend.position = "none",
    panel.border = element_rect(fill = NA, color = "black"),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks = element_line(color = "black", linewidth = 0.4),
    axis.ticks.length.x = unit(-4, "pt"),
    axis.ticks.length.y = unit(-4, "pt"),
    axis.text = element_text(family =     "Arial", size = 10),
    axis.text.x = element_text(family =   "Arial", size = 11, margin = margin(t = 8)),
    axis.text.y = element_text(family =   "Arial", size = 11, margin = margin(r = 8)),
    axis.title = element_text(family =    "Arial", size = 11),
    plot.subtitle = element_text(family = "Arial", size = 10),
    plot.title = element_text(family =    "Arial", size = 12),
    strip.text = element_text(family =    "Arial", size = 11),
    plot.tag = element_text(family =      "Arial", size = 12, face = "bold")
  )
```

## FIGURE 2

```{r}
fig2_theme <- plot_theme +
  theme(
    panel.border = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.4),
    axis.text.y = element_text(family = "Arial", size = 8, margin = margin(r = 8)),
    axis.text.x = element_text(family = "Arial", size = 10, margin = margin(t = 8))
  )


# Habitat & Metro Plot Function
make_city_plot <- function(data, yvar, ylabel, ybreaks, ylimits, emm_city) {
  ggplot(data, aes(x = Metro, y = .data[[yvar]])) +
    geom_boxplot(
      aes(fill = Habitat),
      color = "black", alpha = 0.8, outlier.shape = NA,
      linewidth = 0.4, position = position_dodge(0.8)
    ) +
    geom_point(
      aes(group = Habitat), shape = 21, fill = "gray60",
      position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8),
      alpha = 0.7, size = 1.5, color = "black", stroke = 0.3
    ) +
    geom_point(
      data = emm_city,
      aes(x = Metro, y = emmean, group = Habitat),
      position = position_dodge(0.8),
      shape = 3, size = 2, stroke = 0.8, color = "red", inherit.aes = FALSE
    ) +
    scale_fill_manual(values = habitat_colors) +
    scale_x_discrete(labels = c("New York City" = "NYC", "Philadelphia" = "PHL",
                                 "Richmond" = "RVA")) +
    scale_y_continuous(breaks = ybreaks, limits = ylimits,
                       guide = guide_axis(minor.ticks = TRUE)) +
    labs(y = ylabel, x = NULL) +
    fig2_theme
}

# Habitat Plot Function
make_habitat_plot <- function(data, yvar, ybreaks, ylimits, sig_y, sig_anno, emm_hab) {
  ggplot(data, aes(x = Habitat, y = .data[[yvar]])) +
    geom_boxplot(
      aes(fill = Habitat),
      color = "black", alpha = 0.8, outlier.shape = NA,
      linewidth = 0.4
    ) +
    geom_jitter(shape = 21, fill = "gray60",
                position = position_jitter(0.07),
                alpha = 0.7, size = 1.5, color = "black", stroke = 0.3) +
    geom_point(
      data = emm_hab,
      aes(x = Habitat, y = emmean),
      shape = 3, size = 2, stroke = 0.8, color = "red", inherit.aes = FALSE
    ) +
    geom_signif(
      comparisons = list(c("Urban", "Rural")),
      annotations = sig_anno,
      y_position = sig_y, tip_length = 0.03,
      textsize = 3, size = 0.3, family = "Arial"
    ) +
    scale_fill_manual(values = habitat_colors) +
    scale_x_discrete(limits = c("Urban", "Rural")) +
    scale_y_continuous(breaks = ybreaks, limits = ylimits,
                       guide = guide_axis(minor.ticks = TRUE)) +
    labs(y = NULL, x = NULL) +
    fig2_theme +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
}

# Emmeans Factors
emm_weight_city_env <- as.data.frame(emmeans(Weightlm2,  ~ Metro * Habitat))
emm_weight_env      <- as.data.frame(emmeans(Weightlm2,  ~ Habitat))
emm_BC_city_env     <- as.data.frame(emmeans(BClm,       ~ Metro * Habitat))
emm_BC_env          <- as.data.frame(emmeans(BClm,       ~ Habitat))
emm_femur_city_env  <- as.data.frame(emmeans(femurlm,   ~ Metro * Habitat))
emm_femur_env       <- as.data.frame(emmeans(femurlm,   ~ Habitat))
emm_cent_city_env   <- as.data.frame(emmeans(centroidlm, ~ Metro * Habitat))
emm_cent_env        <- as.data.frame(emmeans(centroidlm, ~ Habitat))
emm_weight_city_env$Habitat <- factor(emm_weight_city_env$Habitat, levels = c("Urban", "Rural"))
emm_weight_env$Habitat      <- factor(emm_weight_env$Habitat,      levels = c("Urban", "Rural"))
emm_BC_city_env$Habitat     <- factor(emm_BC_city_env$Habitat,     levels = c("Urban", "Rural"))
emm_BC_env$Habitat          <- factor(emm_BC_env$Habitat,          levels = c("Urban", "Rural"))
emm_femur_city_env$Habitat  <- factor(emm_femur_city_env$Habitat,  levels = c("Urban", "Rural"))
emm_femur_env$Habitat       <- factor(emm_femur_env$Habitat,       levels = c("Urban", "Rural"))
emm_cent_city_env$Habitat   <- factor(emm_cent_city_env$Habitat,   levels = c("Urban", "Rural"))
emm_cent_env$Habitat        <- factor(emm_cent_env$Habitat,        levels = c("Urban", "Rural"))

weight1 <- make_city_plot(weight, "Weight", "Weight (g)", seq(6, 32, by = 4), c(6, 32), emm_weight_city_env)
weight2 <- make_habitat_plot(weight, "Weight", seq(6, 32, by = 4), c(6, 32), 29.4, "**", emm_weight_env)
weight_combined <- weight1 + weight2 + plot_layout(widths = c(1.2, 1))

BC1 <- make_city_plot(weight, "BC", "Body Condition (g/mm)", seq(0.10, 0.32, by = 0.04), c(0.10, 0.32), emm_BC_city_env)
BC2 <- make_habitat_plot(weight, "BC",seq(0.10, 0.32, by = 0.04), c(0.10, 0.32), 0.30, "**", emm_BC_env)
BC_combined <- BC1 + BC2 + plot_layout(widths = c(1.2, 1))

femur1 <- make_city_plot(linear, "femur", "Femur Length (mm)", seq(9, 18, by = 1), c(9, 18), emm_femur_city_env)
femur2 <- make_habitat_plot(linear, "femur", seq(9, 18, by = 1), c(9, 18), 17, "*", emm_femur_env)
femur_combined <- femur1 + femur2 + plot_layout(widths = c(1.2, 1))

cent1 <- make_city_plot(skull, "centroid", "Centroid Size", seq(86, 116, by = 4), c(86, 116), emm_cent_city_env)
cent2 <- make_habitat_plot(skull, "centroid", seq(86, 117, by = 4), c(86, 117), 114, "*", emm_cent_env)
cent_combined <- cent1 + cent2 + plot_layout(widths = c(1.2, 1))

# Save
ggsave("weight_combined.pdf",  plot = weight_combined,  width = 88, height = 50, units = "mm", dpi = 600, device = cairo_pdf)
ggsave("BC_combined.pdf",      plot = BC_combined,      width = 88, height = 50, units = "mm", dpi = 600, device = cairo_pdf)
ggsave("femur_combined.pdf",   plot = femur_combined,   width = 88, height = 50, units = "mm", dpi = 600, device = cairo_pdf)
ggsave("cent_combined.pdf",    plot = cent_combined,    width = 88, height = 50, units = "mm", dpi = 600, device = cairo_pdf)

#Print
print(weight_combined)
print(BC_combined)
print(femur_combined)
print(cent_combined)
```

## FIGURE 3

```{r}
# PC1 vs PC2 Plot

pc1_pc2_plot <- ggplot(skull_pca, aes(PC1, PC2)) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 3, alpha = 0.7,
             color = "black", stroke = 0.4) +
  stat_ellipse(aes(group = Habitat, color = Habitat), type = "t",
               linetype = "dashed", linewidth = 0.8) +
  scale_fill_manual(values = habitat_colors) +
  scale_color_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_x_continuous(n.breaks = 6, guide = guide_axis(minor.ticks = TRUE)) +
  scale_y_continuous(n.breaks = 6, guide = guide_axis(minor.ticks = TRUE)) +
  coord_fixed() +
  labs(x = "PC1", y = "PC2") +
  plot_theme

pc1_pc2_plot <- pc1_pc2_plot +
  plot_annotation(tag_levels = list(c("A")))

colnames(skull_pca)


skull_pca$log_centroid <- gdf$log_centroid
sum(is.na(skull_pca$log_centroid))
sum(is.na(skull_pca$Habitat))
sum(is.na(skull_pca$Metro))

range(skull_pca$PC1)
range(skull_pca$PC2)

cor_pc1 <- cor.test(skull_pca$log_centroid, skull_pca$PC1)

# (log)Centroid vs PC1 plot
centroid_pc1_plot <- ggplot(skull_pca, aes(x = log_centroid, y = PC1)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 3, alpha = 0.7,
             color = "black", stroke = 0.4) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_x_continuous(n.breaks = 6, guide = guide_axis(minor.ticks = TRUE)) +
  scale_y_continuous(limits = c(-0.042, 0.042), n.breaks = 6,
                     guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "(log) Centroid Size", y = "PC1",
       subtitle = paste0("PC1 ~ (log) Centroid Size: r = ", round(cor_pc1$estimate, 2))) +
  plot_theme

# PC1 Boxplot
pc1_boxplot <- ggplot(skull_pca, aes(x = Habitat, y = PC1)) +
  geom_boxplot(
    aes(fill = Habitat),
    color = "black", alpha = 0.8, outlier.shape = NA,
    linewidth = 0.4
  ) +
  geom_jitter(aes(fill = Habitat, shape = Metro), position = position_jitter(0.07),
              alpha = 0.7, size = 3, color = "black", stroke = 0.4,
              show.legend = FALSE) +
  geom_signif(
    comparisons = list(c("Urban", "Rural")),
    annotations = c("**"),
    textsize = 5, y_position = 0.0345,
    tip_length = 0.02, size = 0.5, color = "black", family = "Arial"
  ) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_x_discrete(limits = c("Urban", "Rural")) +
  scale_y_continuous(limits = c(-0.042, 0.042), n.breaks = 6,
                     guide = guide_axis(minor.ticks = TRUE)) +
  labs(y = "PC1", x = "Habitat") +
  plot_theme

combined_pc1 <- (centroid_pc1_plot | pc1_boxplot) +
  plot_layout(widths = c(2, 1)) +
  plot_annotation(tag_levels = list(c("B", "C")))

# Save
ggsave("pc1_pc2_plot1.pdf", pc1_pc2_plot, width = 165, height = 150, units = "mm", device = cairo_pdf)
ggsave("pc1_combined_with_boxplot.pdf", combined_pc1,  width = 180, height = 120, units = "mm", device = cairo_pdf)

# Print
pc1_pc2_plot
combined_pc1
```

## SUP_FIG_3

```{r}
df_cac <- data.frame(
  CACs    = CAC_scores,
  PC1s    = PC1_scores,
  Habitat = skull_pca$Habitat,
  Metro   = skull_pca$Metro
)

cor_cac <- cor.test(df_cac$CACs, df_cac$PC1s)

sup_fig3 <- ggplot(df_cac, aes(x = CACs, y = PC1s)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 3, alpha = 0.7,
             color = "black", stroke = 0.4) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = metro_shapes) +
  labs(x = "CAC Scores", y = "PC1 Scores",
       subtitle = paste0("r = ",  round(cor_cac$estimate, 3))) +
  plot_theme +
  theme(legend.position = "none")

ggsave("SUP_FIG_3.pdf", sup_fig3,
       width = 100, height = 80, units = "mm", device = cairo_pdf)

print(sup_fig3)
```

## SUP_FIG_4

```{r}
colnames(skull)
skull$Endocast.Volume..mm3.    <- as.numeric(skull$Endocast.Volume..mm3.)
skull_pca$Endocast.Volume..mm3. <- as.numeric(skull_pca$Endocast.Volume..mm3.)
skull$centroid                    <- as.numeric(gdf$centroid)
skull_pca$centroid                <- as.numeric(skull$centroid)
skull_pca$Body.Length..mm.        <- as.numeric(skull_pca$Body.Length..mm.)
skull_pca$Weight..g.              <- as.numeric(skull_pca$Weight..g.)
linear$Body.Length..mm.           <- as.numeric(linear$Body.Length..mm.)
linear$Femur.Length..mm.          <- as.numeric(linear$Femur.Length..mm.)
skull_pca$Femur.Length..mm.       <- as.numeric(skull_pca$Femur.Length..mm.)

# Correlation Tests
cor_weight   <- cor.test(skull_pca$PC1, skull_pca$Weight..g.)
print(cor_weight)
cor_femur    <- cor.test(skull_pca$PC1, skull_pca$Femur.Length..mm.)
print(cor_femur)
cor_bl  <- cor.test(skull_pca$PC1, skull_pca$Body.Length..mm.)
print(cor_bl)
cor_volume   <- cor.test(skull_pca$PC1, skull_pca$Endocast.Volume..mm3.)
print(cor_volume)
cor_cent     <- cor.test(skull$centroid, skull$Endocast.Volume..mm3.)
print(cor_cent)
cor_femur_bl <- cor.test(linear$Body.Length..mm., linear$Femur.Length..mm.)
print(cor_femur_bl)

help("cor.test")

sup_theme <- plot_theme +
  theme(
    panel.border  = element_blank(),
    axis.line     = element_line(color = "black", linewidth = 0.4),
    axis.title    = element_text(family = "Arial", size = 10),
    axis.text     = element_text(family = "Arial", size = 9),
    plot.subtitle = element_text(family = "Arial", size = 8)
  )

legend_guides <- guides(
  fill  = guide_legend(title = "Habitat",
                       override.aes = list(shape = 21, size = 3)),
  shape = guide_legend(title = "Metro",
                       override.aes = list(fill = "gray50", size = 3))
)

p1 <- ggplot(skull_pca, aes(x = PC1, y = Weight..g.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "PC1", y = "Weight (g)",
       subtitle = paste0("r = ", round(cor_weight$estimate, 3))) +
  sup_theme +
  legend_guides

p2 <- ggplot(skull_pca, aes(x = PC1, y = Femur.Length..mm.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "PC1", y = "Femur Length (mm)",
       subtitle = paste0("r = ", round(cor_femur$estimate, 3))) +
  sup_theme +
  legend_guides

p3 <- ggplot(skull_pca, aes(x = PC1, y = Body.Length..mm.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "PC1", y = "Body Length (mm)",
       subtitle = paste0("r = ", round(cor_bl$estimate, 3))) +
  sup_theme +
  legend_guides

p4 <- ggplot(skull_pca, aes(x = PC1, y = Endocast.Volume..mm3.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "PC1", y = "Endocast Volume (mm³)",
       subtitle = paste0("r = ", round(cor_volume$estimate, 3))) +
  sup_theme +
  legend_guides

p5 <- ggplot(skull, aes(x = centroid, y = Endocast.Volume..mm3.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "Centroid Size", y = "Endocast Volume (mm³)",
       subtitle = paste0("r = ", round(cor_cent$estimate, 3))) +
  sup_theme +
  legend_guides

p6 <- ggplot(linear, aes(x = Body.Length..mm., y = Femur.Length..mm.)) +
  geom_smooth(method = "lm", se = TRUE, fill = "gray70", color = "black",
              linewidth = 0.8, alpha = 0.6) +
  geom_point(aes(fill = Habitat, shape = Metro), size = 2, alpha = 0.7,
             color = "black", stroke = 0.3) +
  scale_fill_manual(values = habitat_colors) +
  scale_shape_manual(values = metro_shapes) +
  scale_y_continuous(n.breaks = 5, guide = guide_axis(minor.ticks = TRUE)) +
  labs(x = "Body Length (mm)", y = "Femur Length (mm)",
       subtitle = paste0("r = ", round(cor_femur_bl$estimate, 3))) +
  sup_theme +
  legend_guides

combined_sup4 <- (p1 | p2 | p3) / (p4 | p5 | p6) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.text     = element_text(family = "Arial", size = 10),
    legend.title    = element_text(family = "Arial", size = 10)
  )

# Print
print(combined_sup4)

# Save
ggsave("SUP_FIG_4.pdf", combined_sup4,
       width = 180, height = 120, units = "mm", device = cairo_pdf)
```
